The Atala PRISM Protocol

### 2022-12-16 prism-backend/research/prism-protocol/preliminaries.tex
New and more generic definition of a global PKI.

\todo{Briefly describe algorithms they use.}

\paragraph{Append-Only Bulletin Boards.} %
%\todo{For now, simplified notion as in \cite{acc+20}.}

\paragraph{Digital Signatures.} %
%Maybe build on \IdealFSig from \cite{canetti03}.

\paragraph{Merkle Trees.}

\paragraph{Public Key Infrastructures.} %
% Maybe build on \IdealFCA from \cite{canetti03}.

\paragraph{Ideal functionality specification.} %

@@

\paragraph{Decentralised Identifiers (DIDs).} %
Much of the SSI community is built on top of DIDs, which is a W3C Recommendation
since 2022\footnote{\url{https://www.w3.org/TR/did-core}. Last access, 3rd
  November, 2022.}. Essentially, the DID specification defines how to associate
public keys and additional metadata to unique identifiers, leveraging verifiable
data registries (VDR). Whereas there is no concrete requirement on these VDRs,
the typical choice are blockchains, which leads to the ``Decentralised''
adjective for DIDs. The DID specification describes the syntax that identifiers
must follow, as well as the operations (CRUD: Create, Read, Update, Deactivate)
that any DID-compatible mechanism must implement.

That is, a mechanism following the DID specification must enable any user to be
able to register a unique identifier, associate public key material to it, and
manange it. Thus, it is natural to see such a mechamism as an equivalent to
traditional Public Key Infrastructures (PKIs), where different instantiations of
such a mechanism -- e.g., with different approaches to implement the CRUD
methods required by the DID specification, or based on different VDRs -- can be
seen as a different type of DID-based PKI.

\todo{Maybe expand. Also, describe the difference between DID Document, and DID.
  Also describe the concept of DID controller, essential to prevent manipulation
  of DIDs and their documents.}

>>>>

% \todo{Briefly describe algorithms they use.}

% \paragraph{Append-Only Bulletin Boards.} %
% %\todo{For now, simplified notion as in \cite{acc+20}.}

% \paragraph{Digital Signatures.} %
% %Maybe build on \IdealFSig from \cite{canetti03}.

% \paragraph{Merkle Trees.}

% \paragraph{Public Key Infrastructures.} %
% % Maybe build on \IdealFCA from \cite{canetti03}.

% \subsection{Base Functionalities}
% \label{ssec:did-funcs}

\todo{Mention public delayed outputs.}
\todo{If the functionalities try to fetch some entry from a list, and the entry
  does not exist, then it aborts (assumed for updates).}

\paragraph{Hashes.} %
We model hashes as a global subroutine. No participant needs extra power (such
as programmability or observability), so we use a conventional random oracle --
called strict random oracle in \cite{cdg+18}. For self-containment, we reproduce
here the formulation.

\begin{figure}[ht]
  \begin{framed}
    \begin{minipage}[t]{\textwidth}
      \textrm{Upon receiving a call $\IdealGRO(\cmd,\sid,\arg)$ from some party
        $P$ to run command \cmd, in session \sid, with arguments \arg,
        \IdealGRO operates as follows.} \\
      \textrm{\IdealGRO keeps a list \hashlist of known hashed messages, and
        outputs hashes of length $l \gets l(n)$, for some fixed security
        parameter $n$.}
    \end{minipage}
    \vspace*{0.5em}

    \centering
    \procedure{$\pcif \cmd = \uccmd{HashQuery} \land \arg = \msg$}{%
      \pcif (\msg,h) \notin \hashlist:
      h \getr \bin^{l}; \hashlist \gets \hashlist \cup \lbrace (\msg,h) \rbrace \\
      \ucio{Output}~(\uccmd{HashConfirm},h)
    }
  \end{framed}
\end{figure}

\paragraph{Global clock.} %
Our ideal (global) functionality for a ledger relies on a global clock to model
delays in appending information. We rely on the ideal global clock functionality
given in \cite{kkk21}. In a nutshell, the functionality keeps track of the
parties in the system, and allows them to signal their intention to advance the
clock's time. When all parties agree on it, the clock's time is incremented. See
\figref{fig:gclock}.

\begin{figure}[ht]
  \begin{framed}
    \begin{minipage}[t]{\textwidth}
      \textrm{Upon receiving a call $\IdealGclock(\cmd,\arg)$ from
        some party $P$ to run command \cmd, with arguments \arg,
        \IdealGclock operates as follows.} \\
      \textrm{\IdealGclock maintains a variable $t$ with the current itme,
        a variable $T$ with the list of parties keeping track of the time,
        and a map $A$ which keeps the parties' (dis)agreements to advance the
        clock.}
    \end{minipage}
    \vspace*{0.5em}

    \begin{minipage}[t]{0.40\textwidth}
      \procedure{$\pcif \cmd = \uccmd{Register} \land \arg = \emptyset$}{
        T \gets T \cup \set{P} \\
        \ucio{Output}~(\uccmd{Registered})~\text{to}~P \\
      }
      \procedure{$\pcif \cmd = \uccmd{Update} \land \arg = \emptyset$}{
        A(P) \gets \top \\
        \pcif \forall P \in T, A(P) = \top: \\
        \pcind t \gets t + 1; \forall P \in T, A(P) = \bot \\
        \pcind \ucio{Send}~(\uccmd{Updated})~\text{to}~P        
      }
    \end{minipage}
    \hspace*{1.25em}
    \begin{minipage}[t]{0.60\textwidth}
      \procedure{$\pcif \cmd = \uccmd{Deregister} \land \arg = \emptyset$}{
        T \gets T \setminus \set{P} \\
        \text{Output}~(\uccmd{DeregisteredOk})~\text{to}~P \\
      }
      \procedure{$\pcif \cmd = \uccmd{Read} \land \arg = \emptyset$}{
        \text{Output}~(\uccmd{ReadOk},t)~\text{to}~P
      }
    \end{minipage}
  \end{framed}
  \caption{Ideal subroutine for a global clock, from \cite{kkk21}.}
  \label{fig:gclock}
\end{figure}

\paragraph{$\delta$-delay Ledger.} %
We adopt a ledger notion given in \cite{kkk21}. This is global functionality
that defines an ideal version of a ledger as a shared subroutine, which models
the append-only property of ledgers, and the fact that changes take some delay
until they are applied to the ledger.
%
This functionality accepts sending individual transactions to be added to the
ledger state, fetching (reading) the state of the ledger, and extending the
ledger by appending (blocks of) transactions to it. Interestingly, it gives the
adversary control over when to advance the status of each party, as long as
these advancements ensure the common prefix rule -- where $\ldgState' \prec
\ldgState$ means that $\ldgState'$ is a prefix of \ldgState. It also relies
on yet another global functionality, the global clock, to timestamp incoming
transactions. If a transaction is not appended to an honest party's view before
the given $\delta$ delay, then the functionality aborts.
%
For the sake of readability, we reproduce the functionality in
\figref{fig:gledger}.

\begin{figure}[ht]
  \begin{framed}
    \begin{minipage}[t]{\textwidth}
      \textrm{Upon receiving a call $\IdealGdledger(\cmd,\arg)$ from
        some party $P$ to run command \cmd, with arguments \arg,
        \IdealGdledger operates as follows.} \\
      \textrm{\IdealGdledger maintains a mapping \ldgMap which associates
        parties with their view of the ledger.
        \ldgUtxo is a multiset of unconfirmed transactions.
        \ldgState is the authoritative ledger state
        All mappings and (multi-)sets are initialised to empty values.
      }
    \end{minipage}
    \vspace*{0.5em}

    \begin{minipage}[t]{0.40\textwidth}
      \procedure{$\pcif \cmd = \uccmd{Submit} \land \arg = \ldgTx$}{
        \ucio{Send}~(\uccmd{Read})~\text{to}~\IdealGclock \\
        \ucio{Recv}~(\uccmd{ReadOk},t)~\text{from}~\IdealGclock \\
        \ldgUtxo \gets \ldgUtxo \cup \set{(\ldgTx,t)} \\ 
        \ucio{Output}~(\uccmd{Transaction},\ldgTx,t)~\text{to}~\adv \\
      }
      \procedure{$\pcif \cmd = \uccmd{Extend} \land \arg = \ldgState'$}{
        \pcif P = \adv \land \ldgState' \subseteq \ldgUtxo: \\
        \pcind \ldgUtxo \gets \ldgUtxo \setminus \ldgState' \\
        \pcind \ldgState \gets \ldgState || \ldgState'
      }
    \end{minipage}
    \hspace*{1.25em}
    \begin{minipage}[t]{0.60\textwidth}
      \procedure{$\pcif \cmd = \uccmd{Read} \land \arg =
        \emptyset$}{
        % \pcif P = \adv: data \gets \ldgState \\
        % \pcelse: data \gets \ldgMap(P) \\
        \ucio{Send}~(\uccmd{Read})~\text{to}~\IdealGclock \\
        \ucio{Recv}~(\uccmd{ReadOk},t)~\text{from}~\IdealGclock \\
        \pcif \exists (\ldgTx,t') \in \ldgUtxo~\suchthat~|t-t'| > \delta:
        \pcreturn \bot \\
        \pcelse \pcif \exists (\ldgTx,t') \in \ldgState~\suchthat~ |t-t'|
        > \delta~\land \\
        \pcind \exists P' \in \ldgHonest~\suchthat~(\ldgTx,t') \notin \ldgMap(P'):
        \pcreturn \bot \\        
        \text{Output}~(\uccmd{ReadOk}, \ldgHead(\ldgMap(P)))~\text{to}~P \\
      }
      \procedure{$\pcif \cmd = \uccmd{Advance} \land \arg = (P',\ldgState')$}{
        \pcif P = \adv \land \ldgMap(P') \prec \ldgState' \prec \ldgState:
        \ldgMap(P') \gets \ldgState'
      }
    \end{minipage}
  \end{framed}
  \caption{Ideal subroutine for a global $\delta$-delay ledger, from
    \cite{kkk21}.}
  \label{fig:gledger}
\end{figure}

\paragraph{Digital Signatures.} %

\begin{figure}
  \begin{framed}
    \begin{minipage}[t]{\textwidth}
      \textrm{Upon receiving a call $\IdealFSig(\cmd,\sid,\arg)$ from some party
        $P$ to run command \cmd, in session \sid, with arguments \arg,
        \IdealFSig operates as follows.} \\
      \textrm{\IdealFSig keeps a list \SIG of known signed messages.}      
    \end{minipage}
    \vspace*{0.5em}

    \begin{minipage}[t]{0.5\textwidth}
      \procedure{$\pcif \cmd = \uccmd{KeyGen} \land \arg =
        \emptyset$}{
        \pccomment{\fbox{Key Generation}} \\
        \pcif \sid = (P,\ssid): \\
        \pcind \ucsend~(\uccmd{KeyGen},\sid)~\text{to}~\adv \\
        \pcind \ucrecv~(\uccmd{VerKey},\sid,v)~\text{from}~\adv \\
        \pcind \ucio{Output}~(\uccmd{VerKey},\sid,v)~\text{to}~P \\
        \pcind \text{Store}~(P,v)
      }
      \vspace*{0.25em}
      \procedure{$\pcif \cmd = \uccmd{Verif} \land
        \arg = (\sig,\msg,v')$}{
        \pccomment{\fbox{Verify a message}} \\
        \ucsend~(\uccmd{Verif},\sid,\msg,\sig,v)~\text{to}~\adv \\
        \ucrecv~(\uccmd{Verifd},\sid,\msg,\phi)~\text{from}~\adv \\
        \pcif v' = v \land (\msg,\sig,v,1) \in \SIG: f \gets 1 \\
        \pcelse \pcif v' = v \land P~\text{is not corrupt}~\land \\
        \pcind (\msg,\sig',v,1) \notin \SIG: \\
        \pcind \SIG \gets \SIG \cup \lbrace (\msg,\sig,v,0) \rbrace;
        f \gets 0 \\
        \pcelse \pcif (\msg,\sig,v',f') \in \SIG: f \gets f' \\
        \pcelse: \\
        \pcind \SIG \gets \SIG \cup \lbrace (\msg,\sig,v',\phi) \rbrace;~
        f \gets \phi \\
        \ucio{Output}~(\uccmd{Verifd},\sid,\msg,f)
      }
    \end{minipage}
    \begin{minipage}[t]{0.5\textwidth}
      \procedure{$\pcif \cmd = \uccmd{Sign} \land \arg = \msg$}{
        \pccomment{\fbox{Sign a message}} \\
        \pcif \sid = (P,\ssid) \\
        \pcind \ucsend~(\uccmd{Sign},\sid,\msg)~\text{to}~\adv \\
        \pcind \ucrecv~(\uccmd{Signed},\sid,\msg,\sig)~\text{from}~
        \adv \\
        \pcind \pcif (\msg,\sig,v,0) \in \SIG: \\
        \pcind \pcind \ucio{Output}~(\uccmd{Error}) \\
        \pcind \pcelse: \\
        \pcind \pcind \SIG \gets \SIG \cup (\msg,\sig,v,1) \\
        \pcind \pcind \ucio{Output}~(\uccmd{Signed},\sid,\msg,\sig)
      }
    \end{minipage}
  \end{framed}
  \caption{Ideal signature functionality, \IdealFSig, from \cite{canetti03}.}
  \label{fig:fsig}
\end{figure}

@@

\paragraph{Decentralised Identifiers (DIDs).} %
Much of the SSI community is built on top of DIDs, which is a W3C Recommendation
since 2022\footnote{\url{https://www.w3.org/TR/did-core}. Last access, 3rd
  November, 2022.}. Strictly speaking, a DID is a string that uniquely
identifies information referring to some entity, whereas such information is
the DID document. The DID controller is yet another entity who is allowed to
apply changes to a DID document. In the sequel, for simplicity we assume that
the controller is also the entity that the information in the DID document
refers to, but this is not necessarily the case in general.

Essentially, the DID specification defines how to associate
public keys and additional metadata to unique identifiers, leveraging verifiable
data registries (VDR). Whereas there is no concrete requirement on these VDRs,
the typical choice are blockchains, which leads to the ``Decentralised''
adjective for DIDs. The DID specification describes the syntax that identifiers
must follow, as well as the CRUD (Create, Read, Update, Deactivate) operations
that any DID-compatible mechanism must implement. Concretely:

\begin{description}
\item[Create.] Must specify how controllers create new DIDs.
\item[Read.] Must specify how the contents of a DID can be fetched and
  authenticated.
\item[Update.] Must specify what updates can controllers apply (if any)
  to existing DIDs.
\item[Deactivate.] Must specify if (parts of) existing DIDs can be signaled
  as no longer valid, and how.
\end{description}

That is, a mechanism following the DID specification must enable any user to be
able to register a unique identifier, associate public key material to it, and
manange it. Thus, as we do here, it is natural to see such a mechanism as an
specific type of Public Key Infrastructures (PKIs), where different
approaches to implement the CRUD methods required by the DID specification, or
based on different VDRs -- can be seen as a different type of DID-based PKI.

A final requirement put by the W3C DID specification concerns authorisation.
Namely, it states (in Section 8.2) that candidate DID methods ``must define how
authorisation is performed''. This concretely means that DID methods have to
define the way in which controllers prove to be entitled to perform create,
update, or deactivate operations.
